Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-20 programlogic.tex
+

We describe a program logic (a separation logic) along the lines suggested earlier, where every assertion is relative
to an address space in which it is interpreted, allowing us to define \emph{virtual points-to} assertions that make claims
about memory locations in a particular address space. Virtual addresses, and even virtual points-to assertions, 
are not tagged with their address spaces in any way. Memory access in this logic is validated through the use 
of virtual points-to assertions in pre-conditions, which guarantee that address translations succeed.
This supports rules for updating not only typical data in memory that happens to be subject to address translation, but
manipulation of the page tables themselves via virtual addresses (as demanded by all modern hardware), also
via virtual points-to assertions.
To support specifications that deal with multiple address spaces, our logic incorporates a hybrid-style modality
$[r](P)$ to state that an assertion is true in another (assertion-specified) address space rather than the address space
currently active in hardware, which is not only useful for virtual memory manager invariants but critical to reasoning
about change-of-address-space.
By developing this within the \iris framework, we obtain additional features (e.g., fractional permissions) that allow us to verify
some of the most subtle and technically challenging instruction sequences in an OS kernel (Section \ref{sec:experiment}).

@@

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of \iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another pointwise lifting already in  \iris~\cite{dang2019rustbelt,dang2022compass}. 
  This similarity is real, but the existing lifting does not appear to work with indexed \coq types
  like our \lstinline|word n| as a domain.
}
\lstinline[language=Coq]|Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$|.
This is the (\rocq) type of assertions in our logic.
Most constructs in \iris's base logic are defined with respect to any BI-algebra (of \coq type \lstinline|bi|), so automatically
carry over to our derived logic.
\add{Since \textsf{vProp}s are ``just'' \textsc{Iris} predicates on a choice of page table root,
it may not be obvious what the gain is over simply using \textsc{Iris} directly.
The gain is not expressive power.
Every modal logic can be translated straitforwardly to a logic without modalities
(nearly always a small variation on the \emph{standard translation} from propositional modal logic
to first-order logic~\cite{blackburn2006modal}\footnote{\add{Itself similar to verification condition generation
for weakest precondition calculi and Hoare logics.}}) that can directly address the logic's models ---
but these translations, in addition to making specifications far more verbose,
generally do not preserve the appealing reasoning principles of modal logics.
This becomes most apparent when discussing our proof rules.
}

@@

\paragraph{Register points-to} 
The assertion $\textsf{r}\;\mapsto_{\textsf{r}}\{q\}\;\textsf{rv}$ ensures the ownership of the register $\rg$ containing the register value $\rv$. 
The fraction $\qfrac$ with value 1 asserts the unique ownership of the register mapping, and grants update permission on it, 
otherwise, any value $0 < \qfrac <1$ represents partial ownership granting readonly permission on the mapping.

@@

\subsection{An Overly-Restrictive Definition for Virtual Memory Addressing}
\label{sec:overly-restrictive}
A natural definition for a virtual points-to
asserting that virtual address \textsf{va} points to a value \textsf{v}
would 
% require that in order for a virtual address \textsf{va} to point to a value \textsf{v}, the assertion 
contain
partial ownership of the physical memory involved in the page table walk that would translate \textsf{va} to
its backing physical location --- with locations existentially quantified since a virtual points-to should not assert
\emph{which} locations are accessed in a page table walk, as in Figure \ref{fig:strongvirtualpointsto}.
It asserts the existence of four page table entries, one at each translation level, and via \lstinline|L4_L1_PointsTo|
asserts that the physical page table walk (per Figure \ref{fig:pagetables}) succeeds in reaching the L1 entry,
which points to the page holding the physical memory backing the virtual address, which contains value \textsf{v}.
Most of the definition lives directly in \textsf{vProp}, using the separation logic structure lifted from \iris's \textsf{iProp}.
\looseness=-1

\todo[inline]{This is a section where it's worth explaining the figure in a little more detail, especially as we reuse L4L1PointsTo
in our own definitions later. I know we discussed at some point, but don't the \texttt{l$n$M52} accessors ignore the \texttt{maddr},
and don't the \texttt{l$n$off} accessors ignore the entry arguments? If so we can drop those (simplifying the figure), and annotate 
(e.g., with comments like \lstinline[mathescape]|(*$\textcircled{E}$*)|) each \texttt{l$n$off} in Figure \ref{fig:strongvirtualpointsto}
with a tag matching a new tag on the corresponding offset chunk of the translated virtual addres in Figure \ref{fig:pagetables}\ldots
}
\lstinline|L4_L1_PointsTo| works by
chaining together the entries for each level, using the sequence of table offsets from the address being translated to index
each table level, and using the physical page address embedded in each entry.\footnote{
  The fractions \lstinline|q1| through \lstinline|q4| represent the fractional ownership of each entry based on how many
  word-aligned addresses might need to share the entry ---  $(\frac{1}{512})^n$ for each level $n$.
}
For example, 
  the first level address translation to get the L4 entry (\lstinline|l4e|) 
  uses the masks \textsf{l4M52} with the current \lstinline|cr3| to get the 
  physical address of the start of the L4 table,
  and \textsf{l4off} with the virtual address being translated to compute the correct offset within that table.
  Then at that physical location is the appropriate entry in the L4 table \textsf{l4e}.
  Subsequent levels of the page table walk work similarly.
The statement of these assertions is simplified by the use of our split physical points-to assertions, since
each level of tables is page-sized.\footnote{We do not address superpages and hugepages in this paper.}
This helper definition is also a more explicit \textsf{vProp}, explicitly binding a value for \lstinline|cr3| and using it
to start the translation process.
\looseness=-1

@@

\subsection{Aliasing/Sharing Physical Pages}
  \label{sec:sharingpages}  
  The virtual points-to definition shown in Figure \ref{fig:strongvirtualpointsto} 
  is too strong to specify some operations a virtual memory manager may need to do, such as
  move one level of the page table to a different physical location while preserving all virtual-to-physical mappings. %\footnote{
  %   x86-64 hardware, like other architectures, includes a feature (which we do not formalize assertions for) to
  %   replace an L1 page table address in an L2 entry with a pointer to a \emph{larger} 2MB page (called super-pages), 
  %   or replace an L2 page table address in an L3 entry with a pointer to a 1GB page (called huge-pages).
  % }
  The use of $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}'s
  virtual points-to definition'
   stores knowledge of the page table walk's details with ownership of the backing physical
  memory.
  Updating any of these mappings (e.g., moving the page tables in physical memory, as in coallescing for
  superpages or hugepages)
  would require explicitly collecting all virtual points-to facts that traverse affected entries.
  It is preferable to permit the page tables themselves to be updated independently of the virtual points-to assertions,
  so long as those updates preserve the same virtual-to-physical translations.
  But this is not possible with Figure \ref{fig:strongvirtualpointsto}'s definition, which ties ownership of particular
  pieces of page table memory to the virtual points-to.

@@

  We separate the physical page-table-walk  
  from the virtual pointsto relation, replacing it with ghost state that merely guarantees the address translation would succeed.
  \iris includes a \emph{ghost map} construction which we use to track mappings from
  virtual addresses to the physical addresses they translate to as a piece of ghost state.
  The map includes for each key in the map (i.e.,
  each virtual address) a token $\ghostmaptoken{\gamma}{k}{v}$ whose ownership is required to update that key-value pair in the ghost map
  named $\gamma$. The existence of such a token implies that the actual map $\theta$ tracked by a corresponding $\mathsf{GhostMap}(\gamma,\theta)$
  resource indeed maps $k$ to $v$. These properties are captured by key \iris rules in Figure \ref{fig:ghostmaps}.\footnote{\iris ghost maps lack established notation in the literature,
  but the syntax we use captures the details of \texttt{iris.base\_logic.lib.ghost\_map}.}
  There are other rules, but these two are most important for explaining ghost maps.
  \textsc{GhostMapUpdate} says that ownership of the actual ghost map with ghost name $\gamma$ and map contents $\theta$,
  and a full (non-fractional) token witnessing that $\theta$ maps \textsf{pa} to \textsf{va} permits an update to the ghost map's state,
  changing the map and replacing the token to represent the new value.
  \textsc{GhostMapLookup} allows using the same information to simply conclude that the mapping indicated by the token is true.

  The \emph{virtual memory manager's invariant} ensures that for each $\ghostmaptoken{\gamma}{\vaddr}{\paddr}$ mapping
  in this map, there are \emph{physical} resources sufficient to ensure that address translation for $\vaddr$
  will resolve in the hardware to $\paddr$ --- via $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$.
  \add{This kernel invariant turns out to be a key ingredient in supporting proofs of VMM functionality:
  in Section \ref{sec:experiment} we will see that separating the logical and physical virtual-to-physical mappings
  is what allows stating the global kernel invariants needed for software page traversals, which prior work did not
  (and could not) pursue.}
  \looseness=-1

@@

  % In Iris this is realized by using an authoritative resource algebra: there is a single \emph{authoritative} global copy of the (ghost)
  % map caching virtual-to-physical address translations, and for each entry a read only \emph{partial} ownership of that key-value pair.
  % The resource algebra itself is instantiated as:
  % \[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  ( (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
For clarity we refer to the specific ghost map summarizing virtual-to-physical translations by \\
\centerline{$\mathcal{A}\textsf{bsPTableWalk}(\delta,\theta) \stackrel{\triangle}{=} \mathsf{GhostMap}(\delta,\theta)$}\\
(omitting $\delta$ for brevity when only one is in scope)
and keep this in a per-address-space invariant described shortly.
We then replace the physical traversal
 $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}'s virtual points-to definition
with ownership of the token \ghostmaptoken{\delta}{\vaddr}{\paddr}, %ghost-map ($ \sumwalkabs\vaddr\qfrac\paddr$),
yielding Figure \ref{fig:virtualpointstosharing}'s definition.
This new definition guarantees that the ghost map contains an mapping from the virtual address ($\vaddr$) to a physical address ($\paddr$),
and thus that the per-address-space invariant will contain the physical resources that guarantee the hardware resolves the translation.

@@

We place the authorative ownership of the ghost translation $\mathcal{A}\textsf{PTableWalk}$ in a per-address-space invariant
$\mathcal{I}$\textsf{ASpace} (Figure \ref{fig:peraspaceinvariant}). 
$\mathcal{I}$\textsf{ASpace} alone allows changes to the page tables 
that preserve overall virtual-to-physical translations.
When combined with the
fragment stored in the virtual points-to (Figure \ref{fig:virtualpointstosharing}),
changes in the virtual-to-physical translations become possible without needing to track all virtual points-to assertions.

\todo[inline]{This next paragraph below explains details, but should first explain the big picture: the current address space
is identified by a paddr, so to state the invariant for the address space named by that paddr we need to look up
what invariants should hold, then assert that those invariants do hold.}
In Figure \ref{fig:peraspaceinvariant},
$\theta$ is the logical map from virtual addresses to the physical addresses they should translate to, 
for the \emph{current} address space, corresponding to the currently installed page tables (the ones
indicated by \lstinline|cr3|). In Figure \ref{fig:virtualpointstosharing}, the existentially-quantified $\delta$
is the ghost name for this map (again, for the current address space).
Because our logic is the first to address the existence of multiple address spaces,
it must ensure that $\theta$ actually corresponds to the intended ghost map for the current address space.
This requires looking up the ghost name (i.e., $\delta$) for the current address space,
which is handled by $m$ --- another logical map (tied to ghost state shortly)
from various page table roots to ghost names for the per-address-space ghost maps discussed above.
$\mathcal{I}\textsf{ASpace}(\theta,m)$ then stores the physical resources for the current address space mappings,
ensuring that the current address space is valid (has an entry in $m$ whose corresponding
authoritative map matches $\theta$), and asserting ownership of the appropriate fractional ownership
of the physical page table walk for each virtual-to-physical translation in $\Theta$ (for the current address space).

\todo[inline]{What's not clear in this next paragraph is the relationship between ASpaceLookup in Figure 5 and the lookup part of Figure 3.
I think Figure 5's $m$ and Figure 3's $\delta{s}$ are the same, aren't they? If so we should say so and use the same name.
If not we should explain why they're different.
}
Virtual points-to assertions must also ensure that they are tied to a valid address space, and because they
use ghost map tokens to ensure the virtual-to-physical translation exists in a specific ghost map, they must also
name that map. In Figure \ref{fig:virtualpointstosharing}, this is done with a ghost map token for \emph{another} ghost map,
one shared across the kernel, holding $m$. Our virtual points-to assertion existentially quanfies the address-space-specific
ghost map $\delta$ mapping virtual addresses to physical addresses, and ensures it exists with a token for the distinguished
global (unique) ghost map named by $\delta{s}$.
% the ghost map of virtual-to-physical translations for a particular address space (the ghost name $\delta$ in Figure \ref{fig:virtualpointstosharing}).
% % for each address space must be locateable from a representation of the specific address space:
% % This means we also require a way to locate the map for the current address space from 
% We support lookup by the current root page table pointer stored in \lstinline|cr3|.
\looseness=-1

@@

\subsection{Address-Space Management}
\label{sec:aspacemanagement}
% So far, we have introduced logical abstractions for a single address space, but VMMs
Real VMMs must
 handle more than one address-space.
Doing so requires a way to talk about other address spaces, and means to switch address spaces.
\begin{figure}
\footnotesize
\begin{mathpar}
    \mbox{$[r](\mathsf{P}) \;: \mathsf{vProp} \; \Sigma \; \stackrel{\triangle}{=} \; \lambda \_,\; \mathsf{P}~\mathsf{r} \vspace{0.5em}$}
    \and
    \mbox{$\textsf{Fact P} \stackrel{\triangle}{=} \;\forall \; ,r \; r' \ldotp  \textsf{P r} \dashv\vdash \textsf{P r'}$}
    \and
    \inferrule*{ }{\textsf{Fact}\;[r](\textsf{P}) }
    \and
    \inferrule*{ }{\textsf{Fact}\;(\textsf{r}\;\mapsto_{\textsf{r}}\{q\}\;\textsf{rv})}
    \and
    \inferrule{ }{\textsf{Fact}\;(\textsf{w} \mapsto_{\textsf{p}} \{q\} \textsf{ v} )}
    \and
    \inferrule{ }{(\textsf{P} \vdash \textsf{Q}) \vdash  ([r](\textsf{P}) \vdash  [r](\textsf{Q}))}
    \and
    \inferrule{ }{[r](\textsf{P} \ast \textsf{Q}) \dashv\vdash ([r](\textsf{P}) \ast [r](\textsf{Q})}\vspace{0.5em}
    \and
    \inferrule{ }{[r](\textsf{P}\land \textsf{Q}) \dashv\vdash [r](\textsf{P})\land [r](\textsf{Q})}
    \and
    \inferrule{ }{[r](\textsf{P}\lor \textsf{Q}) \dashv\vdash [r](\textsf{P})\lor [r](\textsf{Q})}
    \and
    \inferrule{ }{\textsf{Fact P} \vdash  [r](\textsf{P}) \dashv\vdash (\textsf{P})} %(\textsf{vProp }\Sigma))\\
\end{mathpar}
\vspace{-1em}
  \caption{Other-space Modality and Its Laws}
  \label{fig:modaldef}
\vspace{-1em}
  \end{figure}
Figure \ref{fig:modaldef} gives the definition of our modal operator for asserting truth of a modal
(address-space-contingent) assertion \emph{in another address space}, which we call
the \emph{other-space} modality. The definition itself is not
particularly surprising --- as our modal assertions are semantically predicates on a page table root (physical)
address, the assertion $[r](P)$ is a modal assertion which ignores the (implicit) current page table root,
and evaluates the truth of $P$ as if $r$ were the page table root. 
The novelty here is not in the details of the definition, but in recognizing that this is the right way to deal with
multiple address spaces, and working out how to support interaction of multiple address spaces (discussed in the next section).
\add{The modal assertions, together with the other-space modality, mean we can give generic definitions
of data structure assertions (e.g., linked lists, etc.) which do not need to track information
about their own address space. In fact, \emph{only} assertions that explicitly deal with multiple address
spaces need to mention address spaces at all (via the other-space modality).
}

We can prove that this modality follows certain basic laws: that its truth is independent of the address
space in which it is considered,  it distributes over various logical connectives, and it follows the rule of
consequence.
We call \textsf{vProp} assertions whose truth is independent of the current address space
\textsf{Fact}s; these include other-space assertions, physical memory points-tos, and register assertions.
\textsf{Fact}s can move in and out of other-space modalities freely.
\looseness=-1

In general, per-address-space invariants should be collected in a larger
VMM invariant, with individual address spaces' invariants pulled out as needed, such as when proving
soundness of an individual virtual memory access.
But such larger invariants would contain many kernel-specific properties that are orthogonal
to the fundamental reasoning principles that are the focus of this paper.
We leave such kernel-specific reasoning to future work, but our verification of task switching
(Section \ref{sec:experiment}) demonstrates support for managing multiple address spaces.
\looseness=-1

>>>>

We describe a program logic (a separation logic) along the lines suggested earlier, where every assertion is relative
to an address space in which it is interpreted, allowing us to define \emph{virtual points-to} assertions that make claims
about memory locations in a particular address space. Virtual addresses, and even virtual points-to assertions, 
are not tagged with their address spaces in any way. Memory access in this logic is validated through the use
of virtual points-to assertions in preconditions, which guarantee that address translations succeed.
This supports rules for updating not only typical data in memory that happens to be subject to address translation but also
manipulation of the page tables themselves via virtual addresses (as demanded by all modern hardware) and also
via virtual points-to assertions.
To support specifications that deal with multiple address spaces, our logic incorporates a hybrid-style modality
$[r](P)$ to state that an assertion is true in another address space (assertion-specified) rather than the address space
currently active in hardware, which is not only useful for virtual memory manager invariants, but also critical to reasoning
about change of address space.
By developing this within the \iris framework, we obtain additional features (e.g., fractional permissions) that allow us to verify
some of the most subtle and technically challenging instruction sequences in an OS kernel (Section \ref{sec:experiment}).

@@

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of \iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another pointwise lifting already in  \iris~\cite{dang2019rustbelt,dang2022compass}. 
  This similarity is real, but the existing lifting does not appear to work with indexed \coq types like our \lstinline|word n| as a domain.
}
\lstinline[language=Coq]|Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$|.
This is the (\rocq) type of assertions in our logic.
Most constructs in \iris's base logic are defined with respect to any BI-algebra (of \coq type \lstinline|bi|), so they automatically
carry over to our derived logic.
\add{Since \textsf{vProp}s are ``just'' \textsc{Iris} predicates on a choice of page table root,
it may not be obvious what the gain is over simply using \textsc{Iris} directly.
The gain is not expressive power.
Every modal logic can be translated straightforwardly to a logic without modalities
(nearly always a small variation on the \emph{standard translation} from propositional modal logic
to first-order logic~\cite{blackburn2006modal}\footnote{\add{Itself similar to verification condition generation
for weakest precondition calculi and Hoare logics.}}) that can directly address the logic's models ---
but these translations, in addition to making specifications far more verbose,
generally do not preserve the appealing reasoning principles of modal logics.
This becomes more apparent when discussing our proof rules.
}

@@

\paragraph{Register points-to} 
The assertion $\textsf{r}\;\mapsto_{\textsf{r}}\{q\}\;\textsf{rv}$ ensures the ownership of the register $\rg$ containing the value of the register $\rv$. 
The fraction $\qfrac$ with value 1 asserts the unique ownership of the register mapping and grants update permission to it;
otherwise, any value $0 < \qfrac <1$ represents partial ownership, granting read-only permission on the mapping.

@@

\subsection{An Overly-Restrictive Definition for Virtual Memory Addressing}
\label{sec:overly-restrictive}
A natural definition for a virtual points-to
asserting that virtual address \textsf{va} points to a value \textsf{v}
would 
% require that in order for a virtual address \textsf{va} to point to a value \textsf{v}, the assertion 
contain
partial ownership of the physical memory involved in the page table walk that would translate \textsf{va} to
its backing physical location --- with locations existentially quantified since a virtual points-to should not assert
\emph{which} locations are accessed in a page table walk, as in Figure \ref{fig:strongvirtualpointsto}.
It asserts the existence of four page-table entries, one at each translation level, and via \lstinline|L4_L1_PointsTo|
asserts that the physical page table walk (per Figure \ref{fig:pagetables}) succeeds in reaching the L1 entry,
which points to the page holding the physical memory backing the virtual address, which contains value \textsf{v}.
Most of the definition lives directly in \textsf{vProp}, using the separation logic structure lifted from \iris's \textsf{iProp}.
\looseness=-1

\todo[inline]{This is a section where it's worth explaining the figure in a little more detail, especially as we reuse L4L1PointsTo
in our own definitions later. I know we discussed at some point, but don't the \texttt{l$n$M52} accessors ignore the \texttt{maddr},
and don't the \texttt{l$n$off} accessors ignore the entry arguments? If so we can drop those (simplifying the figure), and annotate 
(e.g., with comments like \lstinline[mathescape]|(*$\textcircled{E}$*)|) each \texttt{l$n$off} in Figure \ref{fig:strongvirtualpointsto}
with a tag matching a new tag on the corresponding offset chunk of the translated virtual addres in Figure \ref{fig:pagetables}\ldots
}
\lstinline|L4_L1_PointsTo| works by
chaining together the entries for each level, using the sequence of table offsets from the address being translated to index
each table level, and using the physical page address embedded in each entry. \footnote{
  The fractions \lstinline|q1| through \lstinline|q4| represent the fractional ownership of each entry based on how many
  word-aligned addresses might need to share the entry ---  $(\frac{1}{512})^n$ for each level $n$.
}
For example, the first-level address translation to get the L4 entry (\lstinline|l4e|) 
  uses the masks \textsf{l4M52} with the current \lstinline|cr3| to get the physical address of the start of the L4 table
  and \textsf{l4off} with the virtual address being translated to compute the correct offset within that table.
  Then, that physical location is the appropriate entry in the L4 table \textsf{l4e}.
  Subsequent levels of the page table walk work similarly.
The statement of these assertions is simplified by the use of our split physical points-to assertions, since
each level of tables is page-sized. \footnote{We do not address superpages and hugepages in this paper.}
This helper definition is also more explicit in \textsf{vProp} which binds a value to \lstinline|cr3| and uses it to start the translation process.
\looseness=-1

@@

\subsection{Aliasing/Sharing Physical Pages}
  \label{sec:sharingpages}  
  The virtual points-to definition shown in Figure \ref{fig:strongvirtualpointsto} 
  is too strong to specify some operations that a virtual memory manager may need to do, such as move one level of the page table to a different physical location while preserving all virtual-to-physical mappings. %\footnote{
  %   x86-64 hardware, like other architectures, includes a feature (which we do not formalize assertions for) to
  %   replace an L1 page table address in an L2 entry with a pointer to a \emph{larger} 2MB page (called super-pages), 
  %   or replace an L2 page table address in an L3 entry with a pointer to a 1GB page (called huge-pages).
  % }
  The use of $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}'s
  virtual points-to definition stores knowledge of the page table walk details with ownership of the backing physical memory.
  Updating any of these mappings (e.g., moving the page tables in physical memory, as in coalescing for superpages or hugepages)
  would require explicitly collecting all virtual points-to facts that traverse affected entries.
  It is preferable to permit the page tables themselves to be updated independently of the virtual points-to assertions,
  so long as those updates preserve the same virtual-to-physical translations.
  But this is not possible with Figure \ref{fig:strongvirtualpointsto}'s definition, which ties ownership of particular pieces of page table memory to the virtual points-to.

@@

  We separate the physical page-table walk from the virtual points-to relation, replacing it with a ghost state that merely guarantees that the address translation would succeed.
  \iris includes a \emph{ghost map} construction, which we use to track mappings from virtual addresses to the physical addresses they translate to as a piece of ghost state.
  The map includes, for each key in the map (i.e.,
  each virtual address), a token $\ghostmaptoken{\gamma}{k}{v}$ whose ownership is required to update that key-value pair in the ghost map named $\gamma$. The existence of such a token implies that the actual map $\theta$ tracked by a corresponding $\mathsf{GhostMap}(\gamma,\theta)$
  resource indeed maps $k$ to $v$. These properties are captured by key \iris rules in Figure \ref{fig:ghostmaps}.\footnote{\iris ghost maps lack established notation in the literature,
  but the syntax we use captures the details of \texttt{iris.base\_logic.lib.ghost\_map}.}
  There are other rules, but these two are most important for explaining ghost maps.
  \textsc{GhostMapUpdate} says that ownership of the actual ghost map with ghost name $\gamma$ and map contents $\theta$,
  and a full (nonfractional) token witnessing that $\theta$ maps \textsf{pa} to \textsf{va} permits an update to the ghost map's state,
  changing the map and replacing the token to represent the new value.
  \textsc{GhostMapLookup} allows using the same information to simply conclude that the mapping indicated by the token is true.

  The \emph{virtual memory manager's invariant} ensures that for each $\ghostmaptoken{\gamma}{\vaddr}{\paddr}$ mapping in this map, there are \emph{physical} resources sufficient to ensure that the address translation for $\vaddr$
will resolve on the hardware to $\paddr$ --- via $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$.
  \add{This kernel invariant turns out to be a key ingredient in supporting proofs of VMM functionality:
  in Section \ref{sec:experiment} we will see that separating the logical and physical virtual-to-physical mappings
  is what allows stating the global kernel invariants needed for software page traversals, which prior work did not (and could not) pursue.}
  \looseness=-1

@@

  % In Iris this is realized by using an authoritative resource algebra: there is a single \emph{authoritative} global copy of the (ghost)
  % map caching virtual-to-physical address translations, and for each entry a read only \emph{partial} ownership of that key-value pair.
  % The resource algebra itself is instantiated as:
  % \[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  ( (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
For clarity, we refer to the specific ghost map summarizing virtual-to-physical translations by \\
\centerline{$\mathcal{A}\textsf{bsPTableWalk}(\delta,\theta) \stackrel{\triangle}{=} \mathsf{GhostMap}(\delta,\theta)$}\\
(omitting $\delta$ for brevity when only one is in scope)
and keep this in a per-address-space invariant described shortly.
We then replace the physical traversal $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}'s virtual points-to definition
with ownership of the token \ghostmaptoken{\delta}{\vaddr}{\paddr}, %ghost-map ($ \sumwalkabs\vaddr\qfrac\paddr$),
yielding Figure \ref{fig:virtualpointstosharing}'s definition.
This new definition guarantees that the ghost map contains an mapping from the virtual address ($\vaddr$) to a physical address ($\paddr$),
and thus that the per-address-space invariant will contain the physical resources that guarantee that the hardware resolves the translation.

@@

We place the authorative ownership of the ghost translation $\mathcal{A}\textsf{PTableWalk}$ in a per-address-space invariant
$\mathcal{I}$\textsf{ASpace} (Figure \ref{fig:peraspaceinvariant}). 
$\mathcal{I}$\textsf{ASpace} alone allows changes to the page tables
that preserve overall virtual-to-physical translations.
When combined with the
fragment stored in virtual points-to (Figure \ref{fig:virtualpointstosharing}),
changes in the virtual-to-physical translations become possible without the need to track all virtual points-to assertions.

\todo[inline]{This next paragraph below explains details, but should first explain the big picture: the current address space
is identified by a paddr, so to state the invariant for the address space named by that paddr we need to look up
what invariants should hold, then assert that those invariants do hold.}
In Figure \ref{fig:peraspaceinvariant},
$\theta$ is the logical map from virtual addresses to the physical addresses they should translate to, 
for the \emph{current} address space, corresponding to the page tables currently installed (the ones
indicated by \lstinline|cr3|). In Figure \ref{fig:virtualpointstosharing}, the existentially quantified $\delta$
is the ghost name for this map (again, for the current address space).
Because our logic is the first to address the existence of multiple address spaces,
it must ensure that $\theta$ actually corresponds to the intended ghost map for the current address space.
This requires looking up the ghost name (i.e. $\delta$) for the current address space,
which is handled by $m$ --- another logical map (tied to ghost state shortly)
from various page table roots to ghost names for the per-address-space ghost maps discussed above.
$\mathcal{I}\textsf{ASpace}(\theta,m)$ then stores the physical resources for the current address space mappings,
ensuring that the current address space is valid (has an entry in $m$ whose corresponding
authoritative map matches $\theta$), and asserting ownership of the appropriate fractional ownership
of the physical page table walk for each virtual-to-physical translation in $\Theta$ (for the current address space).

\todo[inline]{What's not clear in this next paragraph is the relationship between ASpaceLookup in Figure 5 and the lookup part of Figure 3.
I think Figure 5's $m$ and Figure 3's $\delta{s}$ are the same, aren't they? If so, we should say so and use the same name.
If not we should explain why they're different.
}
Virtual points-to assertions must also ensure that they are tied to a valid address space, and because they
use ghost map tokens to ensure the virtual-to-physical translation exists in a specific ghost map, they must also
name that map. In Figure \ref{fig:virtualpointstosharing}, this is done with a ghost map token for \emph{another} ghost map,
one shared across the kernel, holding $m$. Our virtual points-to assertion existentially quantifies the address-space-specific
ghost map $\delta$ mapping virtual addresses to physical addresses, and ensures it exists with a token for the distinguished
global (unique) ghost map named by $\delta{s}$.
% the ghost map of virtual-to-physical translations for a particular address space (the ghost name $\delta$ in Figure \ref{fig:virtualpointstosharing}).
% % for each address space must be locateable from a representation of the specific address space:
% % This means we also require a way to locate the map for the current address space from 
% We support lookup by the current root page table pointer stored in \lstinline|cr3|.
\looseness=-1

@@

\subsection{Address-Space Management}
\label{sec:aspacemanagement}
% So far, we have introduced logical abstractions for a single address space, but VMMs
Real VMMs must
 handle more than one address-space.
Doing so requires a way to talk about other address spaces, and means to switch address spaces.
\begin{figure}
\footnotesize
\begin{mathpar}
    \mbox{$[r](\mathsf{P}) \;: \mathsf{vProp} \; \Sigma \; \stackrel{\triangle}{=} \; \lambda \_,\; \mathsf{P}~\mathsf{r} \vspace{0.5em}$}
    \and
    \mbox{$\textsf{Fact P} \stackrel{\triangle}{=} \;\forall \; ,r \; r' \ldotp  \textsf{P r} \dashv\vdash \textsf{P r'}$}
    \and
    \inferrule*{ }{\textsf{Fact}\;[r](\textsf{P}) }
    \and
    \inferrule*{ }{\textsf{Fact}\;(\textsf{r}\;\mapsto_{\textsf{r}}\{q\}\;\textsf{rv})}
    \and
    \inferrule{ }{\textsf{Fact}\;(\textsf{w} \mapsto_{\textsf{p}} \{q\} \textsf{ v} )}
    \and
    \inferrule{ }{(\textsf{P} \vdash \textsf{Q}) \vdash  ([r](\textsf{P}) \vdash  [r](\textsf{Q}))}
    \and
    \inferrule{ }{[r](\textsf{P} \ast \textsf{Q}) \dashv\vdash ([r](\textsf{P}) \ast [r](\textsf{Q})}\vspace{0.5em}
    \and
    \inferrule{ }{[r](\textsf{P}\land \textsf{Q}) \dashv\vdash [r](\textsf{P})\land [r](\textsf{Q})}
    \and
    \inferrule{ }{[r](\textsf{P}\lor \textsf{Q}) \dashv\vdash [r](\textsf{P})\lor [r](\textsf{Q})}
    \and
    \inferrule{ }{\textsf{Fact P} \vdash  [r](\textsf{P}) \dashv\vdash (\textsf{P})} %(\textsf{vProp }\Sigma))\\
\end{mathpar}
\vspace{-1em}
  \caption{Other-space Modality and Its Laws}
  \label{fig:modaldef}
\vspace{-1em}
  \end{figure}
Figure \ref{fig:modaldef} gives the definition of our modal operator for asserting the truth of a modal
(address-space-contingent) assertion \emph{in another address space}, which we call
the \emph{other-space} modality. The definition itself is not
particularly surprising --- as our modal assertions are semantic predicates on a page table root (physical)
address, the assertion $[r](P)$ is a modal assertion that ignores the (implicit) current page table root,
and evaluates the truth of $P$ as if $r$ were the page table root. 
The novelty here is not in the details of the definition, but in recognizing that this is the right way to deal with
multiple address spaces, and working out how to support interaction of multiple address spaces (discussed in the next section).
\add{The modal assertions, together with the other-space modality, mean we can give generic definitions
of data structure assertions (e.g., linked lists, etc.) which do not need to track information
about their own address space. In fact, \emph{only} assertions that explicitly deal with multiple address
spaces need to mention address spaces at all (via the other-space modality).
}

We can prove that this modality follows certain basic laws, showing that its truth is independent of the address
space in which it is considered, that it is distributed over various logical connectives, and that it follows the rule of
consequence.
We call \textsf{vProp} assertions whose truth is independent of the current address space
\textsf{Fact}s; these include other address-space assertions, physical memory points-tos, and register assertions.
\textsf{Fact}s can freely move in and out of other address-space modalities.
\looseness=-1

In general, per-address-space invariants should be collected in a larger
VMM invariant, with individual address spaces' invariants pulled out as needed, such as when proving
soundness of an individual virtual memory access.
However, such larger invariants would contain many kernel-specific properties that are orthogonal
to the fundamental reasoning principles that are the focus of this paper.
We leave such kernel-specific reasoning to future work, but our verification of task switching
(Section \ref{sec:experiment}) demonstrates support for managing multiple address spaces.
\looseness=-1
